---
format:
  gfm:
    default-image-extension: ""
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rome

<!-- badges: start -->
[![R-CMD-check](https://github.com/Huber-group-EMBL/rome/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Huber-group-EMBL/rome/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

rome is a minimal R package to reading, writing and validating multiscale [OME-ZARR](https://ngff.openmicroscopy.org/index.html) (or OME-NGFF) images. 

OME-ZARR is a cloud-friendly data format for storing large bioimaging datasets, such as microscopy images, that combines **Zarr**, a chunked, compressed array storage format ([https://zarr.dev/](https://zarr.dev/)) designed for scalable access to multidimensional data, together with **OME-NGFF** ([https://ngff.openmicroscopy.org/](https://ngff.openmicroscopy.org/)) metadata standards for describing multiscale images, segmentations, and coordinate transformations for bioimaging data formats.

The package also provides helpers and methods to manipulate the resulting `ome_zarr` objects in the same way one would manipulate traditional arrays in R. For example, you can subset an `ome_zarr` object using the `[` operator, and the subsetting will be applied to all levels of the multiscale OME-Zarr object.

## Installation

You can install the development version of rome like so:

``` r
# install.packages("pak")
pak::pak("Huber-group-EMBL/rome")
```

## Reading OME-ZARR images

This example shows how to read an OME-ZARR image of version 0.4. 
By default, data are read lazily using `ZarrArray`. 

```{r read}
#| label: read
library(rome)
library(utils)
omezarrzip <- system.file("extdata", "test_ngff_image_v04.ome.zarr.zip", package = "rome")
dir.create(td <- tempfile())
unzip(omezarrzip, exdir = td)
x <- ome_read(td)
plot(x, 1)
```

For remote OME-ZARR files, you can use the `paws.storage::s3` client to read the data directly from the S3 bucket without downloading it first:

```{r read_remote, eval=FALSE}
#| label: read_remote
#| eval: false
library(paws)
s3_client <- paws.storage::s3(
  config = list(
    credentials = list(anonymous = TRUE),
    region = "auto",
    endpoint = "https://uk1s3.embassy.ebi.ac.uk"
  )
)
x <- ome_read(
  "https://uk1s3.embassy.ebi.ac.uk/idr/zarr/v0.4/idr0076A/10501752.zarr", 
  s3_client = s3_client,
)
```

## Writing OME-ZARR images

rome also provides utilities for writing OME-ZARR images for OME-NGFF versions 
0.4 and 0.5. The package also supports writing pyramids using `scalefactors`
argument.

```{r write}
#| label: write
# read image
library(EBImage)
img_file <- system.file("extdata", "example_RGB.png", package="rome")
img <- readImage(img_file)

# write image pyramid
ome_img <- ome_write(img,
                     path = tempfile(fileext = ".ome.zarr"),
                     version = "0.4",
                     scalefactors = c(2,2,3),
                     storage_options = list(chunk_dim = c(64,64,1)))
plot(ome_img)
```