---
title: "Introduction to `romeo`"
date: "`r BiocStyle::doc_date()`"
author:
  - name: Hugo Gruson
    affiliation: EMBL Heidelberg
    email: hugo.gruson@embl.de
  - name: Artür Manukyan
    affiliation: MDC-BIMSB
    email: artur-man@hotmail.com
package: "`r BiocStyle::pkg_ver('romeo')`"
vignette: >
  %\VignetteIndexEntry{Introduction to romeo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document
---

# Introduction

## romeo

`r BiocStyle::Biocpkg("romeo")` is a minimal R package that provides tools to read, validate, and write multiscale images and labels (image labels, segmentation masks, etc.) stored as OME-ZARR files.

The package also provides helpers and methods to manipulate the OME-ZARR images and labels (as `ome_zarr` objects) the same way one would manipulate traditional arrays in R. You can subset an `ome_zarr` object like data arrays (using the `[` operator) where subsetting is applied to all levels of the multiscale OME-Zarr object. 

`r BiocStyle::Biocpkg("romeo")` uses the `r BiocStyle::Biocpkg("Rarr")` package to manipulate images stored as Zarr datasets and OME-ZARR metadata while the `r BiocStyle::Biocpkg("ZarrArray")` package is used to lazily read larger-than-memory images.

## What is OME-ZARR?

OME-ZARR is a cloud-friendly data format for storing large bioimaging datasets, such as microscopy images. It combines: 

* **(i)** **Zarr**, a chunked, compressed array storage format ([https://zarr.dev/](https://zarr.dev/)) designed for scalable access to multidimensional data and
* **(ii)** **OME Next-Generation File Formats**, or **OME-NGFF** ([https://ngff.openmicroscopy.org/](https://ngff.openmicroscopy.org/)), that defines standardized structures and metadata conventions for multiscale labels, segmentations, and coordinate transformations of bioimaging datasets. 

In essence, an OME-ZARR file is a collection of data arrays with XYZCT dimensions representing an image pyramid, combined with metadata (lives in the attributes property of Zarr arrays) that describes the properties of these arrays, such as scales, annotations and coordinate spaces (Figure 1).    

Currently, there exists multiple OME-ZARR formats each having its own OME-NGFF specifications (0.3, 0.4, 0.5 etc.) and Zarr formats (versions 2 or 3). Currently, `r BiocStyle::Biocpkg("romeo")` provides utilities for manipulating OME-ZARR datasets using NGFF versions 0.4 and 0.5.. The current released version of the OME-ZARR specification is 0.5. See [https://ngff.openmicroscopy.org/specifications](https://ngff.openmicroscopy.org/specifications) for more information.

<table>
<tr>
<td>
```{r, echo = FALSE}
system.file("figures", "chunks.svg", package = "romeo") |>
  knitr::include_graphics()
```
</td>
<td>
```{r, out.width = "20%", result = "asis", echo = FALSE}
system.file("figures", "metadata.json", package = "romeo") |>
  readLines() |>
  cat(sep = "\n")
```
</td>
</tr>
</table>
  
# Installation

You can install the development version of `r BiocStyle::Biocpkg("romeo")` like so:

```{r install, eval=FALSE}
install.packages("pak")
pak::pak("Huber-group-EMBL/romeo")
```

# Reading OME-ZARR files

## Images

This is a basic example which shows you how to read an OME-ZARR image of version 0.4. 
By default, data are read lazily using `ZarrArray`. 

```{r read}
library(romeo)
library(utils)
omezarrzip <- system.file("extdata", "test_ngff_image_v04.ome.zarr.zip", package = "romeo")
td <- withr::local_tempfile(fileext = ".ome.zarr")
dir.create(td)
unzip(omezarrzip, exdir = td)
x <- ome_read(td)
plot(x, 1)
```

Alternatively, the data can be read into memory:

```{r read_lazy}
x <- ome_read(td, lazy = FALSE)
```

## Labels

Labels of image pyramids can also be read as images

```{r read_label}
omezarrzip <- system.file("extdata", "test_ngff_image_v04.ome.zarr.zip", package = "romeo")
td <- withr::local_tempfile(fileext = ".ome.zarr")
dir.create(td)
unzip(omezarrzip, exdir = td)
x <- ome_read(file.path(td, "labels/blobs"))
plot(x, all = TRUE)
```

# Reading from S3 storage

For remote OME-ZARR files, you can use the `paws.storage::s3` client to read 
the data directly from the S3 bucket without downloading it first:

```{r read_remote}
library(paws)
s3_client <- paws.storage::s3(
  config = list(
    credentials = list(anonymous = TRUE),
    region = "auto",
    endpoint = "https://uk1s3.embassy.ebi.ac.uk"
  )
)
x <- ome_read(
  "https://uk1s3.embassy.ebi.ac.uk/idr/zarr/v0.4/idr0076A/10501752.zarr",
  s3_client = s3_client,
)
plot(x[1:2, 1:50, 1:50])
```

# Writing OME-ZARR files

## Images

`r BiocStyle::Biocpkg("romeo")` also provides utilities for writing OME-ZARR 
images compatible with OME-NGFF versions 0.4 and 0.5. 

```{r write}
library(EBImage)
img_file <- system.file("extdata", "example_RGB.png", package = "romeo")
img <- readImage(img_file)

# write image pyramid
ome_img <- ome_write(img,
                     path = tempfile(fileext = ".ome.zarr"),
                     axes = c("x", "y", "c"),
                     version = "0.4",
                     storage_options = list(chunk_dim = c(64, 64, 1)))
plot(ome_img, 1)
```

Users can also define their own scaling factors to write image pyramids.
For a `scalefactors` vector with length three, the resulting pyramid will 
contain four scales. Each scale factor in the vector defines the scale 
factor relative to the previous scale.

```{r write2}
ome_img <- ome_write(img,
                     path = tempfile(fileext = ".ome.zarr"),
                     axes = c("x", "y", "c"),
                     version = "0.5",
                     scalefactors = c(2, 2, 3),
                     storage_options = list(chunk_dim = c(64, 64, 1)))
```

## Labels

OME-ZARR label pyramids can be generated in the same way. We first create
our own label data using EBImage. 

```{r image-label}
library(EBImage)

# read the first frame of image
nuc <- readImage(system.file("images", "nuclei.tif", package = "EBImage"))
nuc <- getFrames(nuc)[[1]]

# threshold using otsu's method
nuc_th <- nuc > otsu(nuc)
```

We can now write the label pyramid. The arguments are similar to those used 
for writing images

```{r write_label}
ome_nuc_th <- ome_write(nuc_th,
                        path = tempfile(fileext = ".ome.zarr"),
                        version = "0.4",
                        scalefactors = c(2, 2, 3),
                        storage_options = list(chunk_dim = c(64, 64)),
                        type = "label")
plot(ome_nuc_th, 3)
```

Additional metadata information about labels can be provided using the 
`label_metadata` argument. 

```{r write_label_meta}
# write label, version 0.4
ome_nuc_th <- ome_write(nuc_th,
                        path = tempfile(fileext = ".ome.zarr"),
                        version = "0.4",
                        scalefactors = c(2, 2, 3),
                        storage_options = list(chunk_dim = c(64, 64)),
                        type = "label",
                        label_name = "blobs",
                        label_metadata = list(
                          colors = list(
                            list(`label-value` = 1, rgba = list(255, 255, 255, 255)),
                            list(`label-value` = 2, rgba = list(0, 255, 255, 128))
                          ),
                          properties = list(
                            list(`label-value` = 1, class = "A"),
                            list(`label-value` = 2, class = "B")
                          )
                        ))
```

If the path already includes an image pyramid, then we should define a name 
(e.g. `blobs`) for the label pyramid associated with the image.

```{r write_image_label}
td <- tempfile(fileext = ".ome.zarr")

# write image pyramid
ome_nuc <- ome_write(nuc,
                     path = td,
                     version = "0.4",
                     storage_options = list(chunk_dim = c(64, 64)))

ome_nuc_th <- ome_write(nuc_th,
                        path = td,
                        version = "0.4",
                        scalefactors = c(2, 2, 3),
                        storage_options = list(chunk_dim = c(64, 64)),
                        type = "label",
                        label_name = "blobs")

# plot
layout(matrix(1:2, nrow = 1))
plot(ome_nuc, 3)
plot(ome_nuc_th, 3)
```

# Appendix

## Session info {.unnumbered}
```{r, session-info, echo = FALSE}
sessionInfo()
```
