---
title: "Introduction to `rome`"
date: "`r BiocStyle::doc_date()`"
author:
  - name: Hugo Gruson
    affiliation: EMBL Heidelberg
    email: hugo.gruson@embl.de
  - name: Artür Manukyan
    affiliation: MDC-BIMSB
    email: artur-man@hotmail.com
package: "`r BiocStyle::pkg_ver('rome')`"
vignette: >
  %\VignetteIndexEntry{Introduction to rome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document
---

```{css, echo=FALSE}
.main-container {
  max-width: 1600px;
  margin-left: auto;
  margin-right: auto;
}
```

# Introduction

## OME-ZARR

## rome

`r BiocStyle::Biocpkg("rome")` is a minimal R package to read and write multiscale OME-Zarr files.

It also provides helper and methods to manipulate the resulting `ome_zarr` objects the same way one would manipulate traditional arrays in R. For example, you can subset an `ome_zarr` object using the `[` operator, and the subsetting will be applied to all levels of the multiscale OME-Zarr object.

# Installation

You can install the development version of `r BiocStyle::Biocpkg("rome")` like so:

``` r
# install.packages("pak")
pak::pak("Huber-group-EMBL/rome")
```

# Reading OME-ZARR data

## Images

This is a basic example which shows you how to read a OME-ZARR image of version 0.4. 
By default, the read will be performed lazily using `ZarrArray`. 

```{r read}
library(rome)
library(utils)
omezarrzip <- system.file("extdata", "test_ngff_image_v04.ome.zarr.zip", package = "rome")
dir.create(td <- tempfile())
unzip(omezarrzip, exdir = td)
x <- ome_read(td)
x
```

Otherwise the read can be performed in memory as:

```{r read_lazy}
x <- ome_read(td, lazy = FALSE)
```

## Labels

Labels of image pyramids can also be read as images

```{r read_label}
omezarrzip <- system.file("extdata", "test_ngff_image_v04.ome.zarr.zip", package = "rome")
dir.create(td <- tempfile())
unzip(omezarrzip, exdir = td)
x <- ome_read(file.path(td, "labels/blobs"))
plot(x, all = TRUE)
```

# Reading from S3 storage

For remote OME-ZARR files, you can use the `paws.storage::s3` client to read 
the data directly from the S3 bucket without downloading it first:

```{r read_remote, eval=FALSE}
library(paws)
s3_client <- paws.storage::s3(
  config = list(
    credentials = list(anonymous = TRUE),
    region = "auto",
    endpoint = "https://uk1s3.embassy.ebi.ac.uk"
  )
)
x <- ome_read(
  "https://uk1s3.embassy.ebi.ac.uk/idr/zarr/v0.4/idr0076A/10501752.zarr", 
  s3_client = s3_client,
)
plot(x, all = TRUE)
```

# Writing OME-ZARR data

## Images

`r BiocStyle::Biocpkg("rome")` also provides utilities for writing OME-ZARR images for OME-NGFF versions 
0.4 and 0.5. 

```{r write}
library(EBImage)
img_file <- system.file("extdata", "example_RGB.png", package="rome")
img <- readImage(img_file)

# write image pyramid
ome_img <- ome_write(img,
                     path = tempfile(fileext = ".ome.zarr"),
                     axes = c("x", "y", "c"),
                     version = "0.4",
                     storage_options = list(chunk_dim = c(64,64,1)))
plot(ome_img, 1)
```

Users can also define there own scaling factors for the image pyramids.
For a `scalefactors` with length three, the pyramid will have four scales. 
Eac scale factor in the vector defines the scale factor relative to the previous
scale.

```{r write2}
ome_img <- ome_write(img,
                     path = tempfile(fileext = ".ome.zarr"),
                     axes = c("x", "y", "c"),
                     version = "0.5", 
                     scalefactors = c(2,2,3),
                     storage_options = list(chunk_dim = c(64,64,1)))
```

## Labels

OME-ZARR label pyramids can be generated the same way. We first create
our own label data using EBImage first. 

```{r image-label}
library(EBImage)

# read the first frame of image
nuc <- readImage(system.file("images", "nuclei.tif", package="EBImage"))
nuc <- getFrames(nuc)[[1]]

# threshold using otsu's method
nuc_th = nuc > otsu(nuc)
```

We can now write the label pyramid. Arguments are similar to how images are 
written.

```{r write_label}
ome_nuc_th <- ome_write(nuc_th,
                        path = tempfile(fileext = ".ome.zarr"),
                        version = "0.4",
                        scalefactors = c(2,2,3),
                        storage_options = list(chunk_dim = c(64,64)), 
                        type = "label")
plot(ome_nuc_th, 3)
```

Additional metadata information about labels can be provided using the 
`label_metadata` argument. 

```{r write_label_meta}
# write label, version 0.4
ome_nuc_th <- ome_write(nuc_th,
                        path = tempfile(fileext = ".ome.zarr"),
                        version = "0.4",
                        scalefactors = c(2,2,3),
                        storage_options = list(chunk_dim = c(64,64)), 
                        type = "label", 
                        label_name = "blobs",
                        label_metadata = list(
                          colors = list(
                           list(`label-value` = 1, rgba = list(255, 255, 255, 255)),
                           list(`label-value` = 2, rgba = list(0, 255, 255, 128)) 
                          ),
                          properties = list(
                           list(`label-value` = 1, class = "A"),
                           list(`label-value` = 2, class = "B")
                          )
                        ))
```

If the path already includes an image pyramid, then we should define a name 
(e.g. `blobs`) for the label pyramid associated with the image.

```{r write_image_label}
td <- tempfile(fileext = ".ome.zarr")

# write image pyramid
ome_nuc <- ome_write(nuc,
                     path = td,
                     version = "0.4",
                     storage_options = list(chunk_dim = c(64,64)))

ome_nuc_th <- ome_write(nuc_th,
                        path = td,
                        version = "0.4",
                        scalefactors = c(2,2,3),
                        storage_options = list(chunk_dim = c(64,64)), 
                        type = "label", 
                        label_name = "blobs")

# plot
layout(matrix(1:2, nrow=1))
plot(ome_nuc, 3)
plot(ome_nuc_th, 3)
```

# Appendix

## Session info {.unnumbered}
```{r, session-info, echo = FALSE}
sessionInfo()
```
